---
title: "Bayesian credible intervals"
runtime: shiny
output: statsr:::statswithr_lab
---

```{r echo=FALSE}
suppressMessages(library(magrittr))
```

<div id="instructions">
Complete all **Exercises**, and submit answers to **Questions** on the Coursera platform.
</div>

## Getting Started

### Load packages

In this lab we will explore some basic Bayesian inference using conjugate priors and credible intervals
to examine some categorical and count data from the [CDC's Behavioral Risk Factor Surveillance System](http://www.cdc.gov/brfss/) (BRFSS). A subset of these data from 2013 have been made available in the `statsr` package, as usual we will first load the package and then the data set. 

Let's load the package,

```{r load-packages, message=FALSE}
library(statsr)
data(brfss)
```

This data set contains 5000 observations of 6 variables

variable         | description
---------------- | ---------------------------------------------
`weight`         |  Respondent's weight in pounds.
`height`         |  Respondent's height in inches.
`sex`            |  Respondent's sex
`exercise`       |  Has the respondent exercised in the last 30 days
`fruit_per_day`  |  How many servings of fruit does the respondent consume per day.
`vege_per_day`   |  How many servings of dark green vegetables does the respondent consume per day.


## Credible Interval Calculator

Below is an interactive tool for visualizing a posterior distribution and credible interval for specific values of the posterior distribution's parameters. We will be using this to explore how both our choice of prior as well as our data affect the distribution of the posterior and the credible interval for our parameter of interest.

Note that this app assumes that you know the posterior distribution as well as the parameters of this distribution. In the remainder of the lab we will walk you through calculating the posterior distribution for the Beta-Binomial and the Gamma-Poisson cases based on real data from the BRFSS, and then you will be asked to come back to this app to finally calculate the credible interval for the parameter of interest. But first, some exercises to get you used to using this app.

```{r}
credible_interval_app()
```

1. Suppose the posterior distribution of $\mu$ follows a Normal distribution with mean 10 and variance 5. Which of the following are the bounds of a 95% credible interval for $\mu$? Answer this question using the app.
<ol>
<li> (-1.96, 1.96) </li>
<li> (0.419, 0.872) </li>
<li> (0.959, 3.417) </li>
<li> (5.618, 14.382) **- CORRECT ANSWER** </li>
</ol>

<div id="exercise">
**Exercise**: Confirm your answer by running the code given below the distribution plot in the app.
</div>
```{r normal-ci}
# type your code for the Exercise here, and Knit

```


2. Suppose the posterior distribution of $p$ follows a Beta distribution with $\alpha = 2$ and $\beta = 5$. Which of the following are the bounds of a 90% credible interval for $p$? Answer this question using the app.
<ol>
<li> (-1.678, 5.678) </li>
<li> (0.043, 0.641) </li>
<li> (0.063, 0.582) **- CORRECT ANSWER** </li>
<li> (0.071, 0.949) </li>
</ol>

<div id="exercise">
**Exercise**: Confirm your answer by running the code given below the distribution plot in the app.
</div>
```{r beta-ci}
# type your code for the Exercise here, and Knit

```


3. Suppose the posterior distribution of $\lambda$ follows a Gamma distribution with $\alpha = 4$ and $\beta = 8$. Which of the following are the bounds of a 99% credible interval for $\lambda$? Answer this question using the app.
<ol>
<li> (-3.284, 11.284) </li>
<li> (0.069, 0.693) </li>
<li> (0.084, 1.372) **- CORRECT ANSWER** </li>
<li> (0.171, 0.969) </li>
</ol>

<div id="exercise">
**Exercise**: Confirm your answer by running the code given below the distribution plot in the app.
</div>
```{r gamma-ci}
# type your code for the Exercise here, and Knit

```


## Beta-Binomial

As we discussed in the videos, the Beta distribution is conjugate to the Binomial distribution - meaning that if we use a Beta prior for the parameter $p$ of the Binomial distribution then the posterior distribution of $p$ after observing data will be another Beta distribution. 

$$ \pi(p) ~\sim~ \text{Beta}(a,b) $$
$$ x\,|\,n,p ~\sim~ \text{Binom}(n,p) $$
$$ p \,|\, x,n ~\sim~ \text{Beta}(\alpha,\beta).$$

Our goal with inference in general is to take specific observations (data) and use them to make useful statements about unknown population parameters of interest. The Beta-Binomial is a Bayesian approach for inference about a single population proportion $p$. Whereas with the frequentist approach we used $\hat{p} = x / n$ we will now just use $x$ and $n$ directly with $x$ being the number of successes obtained from $n$ identical Bernoulli trials. As such, we can view $x$ as a Binomial random variable with $n$ trials and $p$ probability of success.
 
To complete our Bayesian specification, all that is left is to define our prior beliefs for the value of $p$ by defining a prior distribution. Our choice of the prior hyperparameters ($a$ and $b$) should reflect our prior beliefs about our parameter of interest. Note that the term hyperparameter defines parameters of prior distributions, and we use the term parameter for defining the unknown parameters of the likelihood. For most conjugate distributions there is usually a straight forward interpretation of these hyperparameters as previously observed data -- in the case of the Beta-Binomial we can think of our hyperparameters as representing $a-1$ previous successes and $b-1$ previous failures.

### Data and the updating rule

Lets start by performing inference on the sex ratio of respondents to the BRFSS, we can define success as being `Female` and we would like to make some statement about the overall sex ratio of American adults based on our sample from the BRFSS. 

```{r}
table(brfss$sex)
```

Let's store the relevant data:
```{r}
n <- length(brfss$sex)
x <- sum(brfss$sex == "Female")
```

For each observed data point from a Binomial ($n$ and $x$) we can calculate the values of the posterior parameters using the following updating rule:

$$ \alpha = a + x $$
$$ \beta = b + n - x $$

From the data we now have $x = `r x`$ (the number of females), and $n - x = `r n - x`$ (the number of males). We'll start with a Beta prior where $a = 1$ and $b = 1$. Remember that this is equivalent to a Uniform distribution. By combining the data with the prior, we arrive at a posterior where 

$$ p \,|\, x,n ~\sim~ \text{Beta}(\alpha = 1 + `r x`,~ \beta = 1 + `r n - x`) $$

4. What is the 95% credible interval for $p$, the proportion of females in the population, based on the posterior distribution obtained with the updating rule shown above. Use the credible interval app to answer this question.
<ol>
<li> some choices
</ol>

5. Which of the following is the correct Bayesian interpretation of this interval?
<ol>
<li> The probability that the true proportion of females lies in this interval is 0.95. </li>
<li> The probability that the true proportion of females lies in this interval is either 0 or 1. </li>
<li> some other choices </li>
</ol>

## Gamma-Poisson

The Gamma-Poisson is another example of a conjugate family where using the Gamma distribution as a prior for the rate parameter $\lambda$ from the Poisson distribution results in a Gamma posterior for $\lambda$.

$$ \pi(\lambda) ~\sim~ \text{Gamma}(a,b) $$
$$ x\,|\,\lambda ~\sim~ \text{Pois}(\lambda) $$
$$ \lambda \,|\, x ~\sim~ \text{Gamma}(\alpha,\beta).$$

For each observed data point from the Poisson ($x$) we can calculate the values of the posterior parameters using the following updating rule:

$$ \alpha = a + x $$
$$ \beta = b + n $$

Once again, our choice of the prior parameters ($a$ and $b$) should reflect our prior beliefs about our parameter of interest ($\lambda$), and in the case of the Gamma-Poisson we can view $a$ as the number of total counts and $b$ as the number of trials.


